---
title: "6: Part 2 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (ANOVA) to real datasets
3. Interpret and report the results of ANOVA in publication-style formats

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
install.packages("htmltools")
install.packages("agricolae")

library(agricolae)   #you may need to install/update package "htmltools" before installing and loading agricolae

Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is used when you have one categorical explanatory variable with more than two categories. In R, we can run the model with the function `lm` or `aov`, the latter of which will allow us to run post-hoc tests to determine pairwise differences.
-have one categorical explanatory variable with more than two levels


Does total litter biomass differ among sites at the Niwot Ridge LTER?
-so we want to check if the mean of litter biomass is diff for different sites in out data set
```{r}
#-remember, one way anova is when you have one categorical variable with more than two levels!!!
#when you run an anova, you reject or don't reject the null. must run post hoc test to see which means aren't equal
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))
#grouping by a few things , what does grouping by mean?

summary(Litter.Totals)
#check all value of plotID
#we want to know why the mean has a high variance!
summary(Litter.Totals$plotID)
#shows number of occurences of each site. Shows us that this isn't a balanced experiment because certain sites have more obs, but can still run the anova

# Test for normality. will check for each of the site locations
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
# shapiro.test() performs the Shapiro-Wilk test of the null that data follows a normal distribution
#shapiro test tests the null hyp that the data follows a normal dist

shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
#we filtered data by site in these equations
#due to the small p value, we reject the null - so it doesn't follow a normal dist
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
#dont reject for the above
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
#dont reject for the above
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])


#results: reject null in all except two: NIWO_057 and NIWO_046
#so those are the only sites that follow normal dist, but we can still use anova!
#qq plots also provide info on normality
qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass)
#the qq plot indicates the data doesn't follow a normal dist!

# Test for equal variance for all the populations
# ANOVA is robust against departures from equal variance.
# bartlett.test() performs Bartlett's test of the null that the variances in each of the groups (samples) are the same.
#can still run anova if results are false
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)
#level is site location
#results: reject null i.e. variances are not equal (p value is too low)

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
#y ~ x
summary(Litter.Totals.anova)
#we have 11 df
#sum square is somethin like deviation from mean of each obs?
#mean square is sum square divided by df
#we reject the null which stated that the mean is the same across all diff sites
#but we don't know which means are diff, so must run post hoc tests

#results: reject null hypothesis i.e. difference between a pair of group means is statiscally significant

# Format ANOVA as lm
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)
#can also run anova tests using the lm function
#lm summary is a little more detailed than the aov one
#all the estimate std are with respect to the int term (which i beleive is the y intercept)

# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
plot(Litter.Totals.anova2)

# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(Litter.Totals.anova)
#the arg here is the result of anova test
#gave us pair wise comparison

# Extract groupings for pairwise relationships
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
#(anova ob, category, group = TRUE (that means we want the function to group levels that have the same mean))
Litter.Totals.groups
#HSD test tries to find groups across the diff levels with sim mean
#same letter indicates similar means
# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)
#will get boxplot for each site id
#for all sites that have a, mean isn't that far from each other
```

What information does the output give us? How might we report this information in a report?

> Want to reorder these sites so that a groups and b groups are grouped together? So we can see changes in the level of the mean

## TWO-WAY ANOVA
### Main effects
A two-way ANOVA allows us to examine the effects of two categorical explanatory variables on a continuous response variable. What if we wanted to know how litter masses varied across functional groups and land cover class?
-remember two way anova allows us to examine the effect of 2 categorical variables on a continuous response variable

```{r}
# Format as aov
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)
#df is associated with # of levels

# Format as lm
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way)
#shows diff bw the means of the groups, lwr + upper interval


```

### Interaction effects
We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an interaction effects two-way ANOVA, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.

```{r}
# Format as aov
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
#we now use the multiply symbol because we want to see the interaction bw them
#the p value indicates that the interaction of the variables is significant
summary(Litter.anova.2way3)

# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.
```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3)

Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass))
#i think we added some information to litter (a data frame we already had), but not totally sure....
#lutter interaction is a combination of land cover and functional group
View(Litter.interaction)
#litter interaction is actually a factor

Litter.test <- cbind(Litter,Litter.interaction)

Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction)

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups

# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = nlcdClass, color = functionalGroup)) +
  geom_boxplot() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
                label = c("d", "d", "d", "a","a", "d", "d", "d","d", "d", "d", "b","d", "d", "d", "d","d", "d", "cd", "bc", "d", "d", "d", "d")) +
  labs(x = "Ecosystem Type", y = "Dry Mass")
print(Litter.anova.plot)




Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
  
  
print(Litter.Totals.plot)
#non parametric means were aren't considering the normal dist stuff, but were not covering those in the vids
#with environmnetal and energy data, we can usually assume normality and move on
#she is happy to point out additional reading to learn non parametric ones



#what does grouping by mean?
#what is a factor? and what are their levels?
#what does the 'with' function do?
#how do we make it so the labels aren't all stacked on top of each other?
```

