---
title: "Assignment 2: Coding Basics"
author: "Courtney Horn"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
counting_by_four <- seq(1,100,4)
#In the code above, I am calculating a sequence of numbers from one to 100, increasing by fours. I also assigned the sequence to an object named "counting_by_four".

#2. 
mean(counting_by_four)
median(counting_by_four)
#I am calculating the mean and median of the sequence I created.

#3. 
mean(counting_by_four) > median(counting_by_four)
#I asked R whether the mean is greater than the median.
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5
setwd("/Users/courtneyannehorn/Desktop/EDA/EDAfin")
  #a)
students_vector <- c("Courtney","Taylor","Lollie", "Bear") # character vector
students
  #b)
testscore_vector <- c(70,90,85,62) # numeric vector
testscore_vector
  #c)
passed_function <- function(x){
ifelse(x>50, TRUE, FALSE) #log_exp, if TRUE, if FALSE
}
passed_function(testscore_vector)
passed_vector <- passed_function(testscore_vector)
passed_vector

#6
students_vector_char <- students_vector
class(students_vector_char)
#students_vector_char is a character vector 

testscore_vector_num <- testscore_vector
class(testscore_vector_num)
#testscore_vector_num is a numerical vector

passed_vector_log <-passed_vector
class(passed_vector_log)
#passed_vector_log is a logical vector



#7 and #8
testperf_df <- data.frame("student" = students_vector_char, "score"=testscore_vector,"pass/nopass"=passed_vector_log)
testperf_df


```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame consists of vectors of different classes (a character vector, a numerical vector, and a logical vector). Matrixes consist of vectors of the same class. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10
passed_function <- function(x){
ifelse(x>50, TRUE, FALSE) #log_exp, if TRUE, if FALSE
}
passed_function(40)

who_passed_results <- function(x){
ifelse(x>50, TRUE, FALSE) #log_exp, if TRUE, if FALSE
}

#11
who_passed_results(testscore_vector)
who_passed_results
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'Ifelse' worked. Any of these would have worked. 'Ifelse' is a way to combine 'if' and 'else', and use fewer lines of coding to accomplish the same task. 
