---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Courtney Horn"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup, message=FALSE}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
library(ggplot2)
#library(ggplot)
NTLRcp <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
colnames(NTLRcp)
class(NTLRcp$sampledate)
head(NTLRcp$sampledate)
#%y is two digit year
#%d is date as number
#%m is month as number


NTLRcp$sampledate <-as.character(NTLRcp$sampledate)
head(NTLRcp$sampledate)
NTLRcp$sampledate <- as.Date(NTLRcp$sampledate, format = "%m/%d/%y")
head(NTLRcp$sampledate)
class(NTLRcp$sampledate)
#NTLRcp


#2
#View(NTLRcp)
#going to build a ggplot for depth by temperature!
#I will use a scatterplot to plot the relationship between two categorical variables
#colnames(NTLRcp)
#tempvsdepth <- 
  #ggplot(NTLRcp, aes(x = depth, y = temperature_C)) +
  #geom_point() +
  #labs(x = "Depth", y = "Temperature (C)") +
#tempvsdepth


mytheme1 <- theme_light(base_size = 12) +
  theme(panel.grid.major = element_line(colour = "black")) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme1)

mytheme2 <- theme_light(base_size = 12) +
  theme(panel.grid.major = element_line(colour = "black")) +
  theme(axis.text = element_text(color = "black"), 
  legend.position = c("right, center"))

theme_set(mytheme2)
#trying to see how my theme changes the plot
tempvsdepth1 <- 
  ggplot(NTLRcp, aes(x = depth, y = temperature_C)) +
  geom_point() +
  labs(x = "Depth", y = "Temperature (C)")
print(tempvsdepth1)




```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The July means of all of the lakes are equal.
Ha: The July means of the lakes are not all equal. At least one of the July lake means is not equal to the others. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, message=FALSE}
#4
#colnames(NTLRcp)

NTLRmonth <- 
  NTLRcp %>% 
  mutate(month = month(sampledate))
colnames(NTLRmonth)
class(NTLRmonth$month)
#NTLRmonth <-as.factor(NTLRmonth)


NTLRcp_wrang <- 
  NTLRmonth %>% 
  filter(month == 7) %>% 
  select(lakename:daynum, depth:temperature_C) %>% 
  drop_na(temperature_C)
#dim(NTLRcp_wrang)
#View(NTLRcp_wrang)

#use NTLRcp_wrang for next questions
#colnames(NTLRcp_wrang)




#5

tempvsdepth3 <- ggplot(data = NTLRcp_wrang, aes(x = depth, y = temperature_C))  +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") +
  ylim(0,35) +
  labs(x = "Depth", y = "Temperature (C)")
tempvsdepth3
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure indicates that temperature decreases linearly as depth increases. Therefore, there is a negative linear relationship between temperature and depth. The distribution of points suggest that the relationship departs from linearity as depth increases. The relationship seems linear between depths of 0 and 7, but doesn't seem to be linear between depths 7 and 15. The temperatures seem consistent within the depth range of 7 - 15.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression, message=FALSE}
#7
tempvsdepth_lm <- lm(data = NTLRcp_wrang, temperature_C ~ depth)
summary(tempvsdepth_lm)
unique(NTLRcp_wrang$depth)
#since there are 27 levels within the explanatory categorical variable, the degrees of freedom is 25 

#summary plots
#plot(tempvsdepth_lm)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 73.8% of the variability in temperature is explained by depth (the adjusted R^2 value = 0.7387). This result is statistically significant (the p value < 2.2e-16). The model predicts that temperature will decrease by 1.95 degrees for each 1m change in depth. There are 25 degrees of freedom for this regression, because there are 27 unique values of depth (the explanatory variable).




---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model, message=FALSE}
#9
#colnames(NTLRcp_wrang)
NTLR_full_lm <- lm(data = NTLRcp_wrang, temperature_C ~ depth + daynum + year4)
summary(NTLR_full_lm)
step(NTLR_full_lm)
#the AIC indicated that the best set of explanatory variables for this purpose includes depth, daynum, and year
AIC(NTLR_full_lm)
NTLR_lm_red1 <-lm(data = NTLRcp_wrang, temperature_C ~ depth + daynum)
AIC(NTLR_lm_red1)
#when I reduced the explanatory variables, the AIC went up (not good)

#10
NTLR_mult_lm <- lm(data = NTLRcp_wrang, temperature_C ~ depth + daynum + year4)
summary(NTLR_mult_lm)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of variables includes depth, daynum, and year. This model explains 74.1% of the observed variance (adjusted R2 = 0.7411). This is an improvement over the model in which the only explanatory variable was depth. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model, message=FALSE}
#12
#one way anova
#colnames(NTLRcp_wrang)
lake_temps_aov <- aov(data = NTLRcp_wrang, temperature_C ~ lakename)
summary(lake_temps_aov)


#unique(NTLRcp$lakename)


#format anova as LM
lake_temps_aov2 <- lm(data = NTLRcp_wrang, temperature_C ~ lakename)
summary(lake_temps_aov2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperature among the lakes (the p value of the anova is < 2.2e-16).



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, message=FALSE}
#14.
tempvsdepth4 <- ggplot(data = NTLRcp_wrang, aes(x = depth, y = temperature_C, color = lakename))  +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  ylim(0,35) +
  labs(x = "Depth", y = "Temperature (C)") +
  theme(legend.position = "bottom")
tempvsdepth4



```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test, message=FALSE}
#15
TukeyHSD(lake_temps_aov)


unique(NTLRcp_wrang$lakename)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: 
Paul Lake and Ward Lake have the same mean temperature (statistically speaking) as Peter Lake. None of the lakes have mean temperatures that are significantly different from all of the other lakes. 


 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 
>Answer: We could use a T-Test to see whether Peter Lake and Paul Lake have distinct mean temperatures. 


